---
title: "Problem set 1"
author: "YOUR NAME HERE"
date: "DATE HERE"
---

# Details

- Who did you collaborate with: TYPE NAMES HERE
- Approximately how much time did you spend on this problem set: ANSWER HERE
- What, if anything, gave you the most trouble: ANSWER HERE


# Task 1

Tell me that you watched the short videos and completed the series of R Markdown tutorials.


# Task 2

## Life expectancy in 2007

```{r load-packages, warning=FALSE, message=FALSE}
# Notice the warning=FALSE and message=FALSE in the chunk options. R spits out 
# a lot of messages when you load tidyverse and we don't want those in the 
# final document.

library(tidyverse)  # This loads ggplot2, dplyr, and other packages you'll need
library(gapminder)  # This loads the gapminder data
```

Let's first look at the first few rows of data:

```{r view-data}
head(gapminder)
```

Right now, the `gapminder` data frame contains rows for all years for all countries. We want to only look at 2007, so we create a new data frame that filters only rows for 2007.

Note how there's a weird sequence of characters: `%>%`. This is called a *pipe* and lets you chain functions together. We could have also written this as `gapminder_2007 <- filter(gapminder, year == 2007)`.

```{r filter-2007}
gapminder_2007 <- gapminder %>%
  filter(year == 2007)

head(gapminder_2007)
```

Now we can plot a histogram of 2007 life expectancies with the default settings:

```{r plot-2007-1}
ggplot(data = gapminder_2007, 
       mapping = aes(x = lifeExp,)) +stat_bin(bins =2)
  geom_histogram()
```

R will use 30 histogram bins by default, but that's not always appropriate, and it will yell at you for doing so. **Adjust the number of bins to 2, then 40, then 100.** **What's a good number for this data? Why?**

TYPE YOUR ANSWER HERE.Every variable has its own pattern of variation, which can reveal interesting information. The best way to understand that pattern is to visualise the distribution of the variable’s values. We can see from the histogram on the Bin=100, the bin width is too small because it shows too much individual data and does not allow the underlying pattern (frequency distribution) of the data to be easily seen. So, from provided 3 bin choices I would pick Bin=2, as in that case I could easily detect general pattern with regard to distribution of data.

```{r plot-2007-2}
ggplot(data = gapminder_2007, 
       mapping = aes(x = lifeExp)) +
  geom_histogram(bins = 2)
```

## Average life expectancy in 2007 by continent

We're also interested in the differences of life expectancy across continents. First, we can group all rows by continent and calculate the mean

This is where the `%>%` function is actually super useful. Remember that it lets you chain functions together—this means we can read these commands as a set of instructions: take the `gapminder` data frame, filter it, group it by continent, and summarize each group by calculating the mean. Without using the `%>%`, we could write this same chain like this: `summarize(group_by(filter(gapminder, year == 2007), continent), avg_life_exp = mean(lifeExp))`. But that's *awful* and impossible to read and full of parentheses that can easily be mismatched.

```{r calc-mean}
gapminder_cont_2007 <- gapminder %>%
  filter(year == 2007) %>% 
  group_by(continent) %>%
  summarize(avg_life_exp = mean(lifeExp))

head(gapminder_cont_2007)
```

Let's plot these averages as a bar chart:

```{r plot-2007-bar}
ggplot(data = gapminder_cont_2007, 
       mapping = aes(x = continent, y = avg_life_exp, fill = continent)) + 
  geom_col()
```

Then, let's plot them as density distributions. We don't need to use the summarized data frame for this, just the original filtered `gapminder_2007` data frame:

```{r plot-2007-density}
ggplot(data = gapminder_2007, 
       mapping = aes(x = lifeExp, fill = continent)) + 
  geom_density()
```

Now let's plot life expectancies as violin charts. These are the density distributions turned sideways:

```{r plot-2007-violin}
ggplot(data = gapminder_2007, 
       mapping = aes(x = continent, y = lifeExp, fill = continent)) + 
  geom_violin()
```

Finally, we can add actual points of data for each country to the violin chart:

```{r plot-2007-violin-points}
ggplot(data = gapminder_2007, 
       mapping = aes(x = continent, y = lifeExp, fill = continent)) + 
  geom_violin() +
  geom_point()
```

The bar chart, density plot, violin plot, and violin plot + points each show different ways of looking at a single number—the average life expectancy in each continent. **Answer these questions:**

- Which plot is most helpful? TYPE YOUR ANSWER HERE: First of all, it  really depends from what perspective you are looking at charts. It really depends what the presenter wants to highlight in his illustration (that is to say what insight he wants to deliver). I would choose bar chart in case "single number—the average life expectancy in each continent"" is the prominent  point of interest.

- Which ones show variability? TYPE YOUR ANSWER HERE:density plot, violin plot, and violin plot + points show variability as they show not only general picture but also behaviour of dataset.They give more details 

- What's going on with Oceania? TYPE YOUR ANSWER HERE.


# Task 3: R and ggplot

```{r load-libraries-1, warning=FALSE, message=FALSE}
# Technically you don't need to run this because we loaded tidyverse earlier in Task 2
library(tidyverse)
```

## 3.2.4

**1: Run `ggplot(data = mpg)`. What do you see?**

```{r blank-plot, fig.width=2, fig.height=2}
ggplot(data = mpg)
```

TYPE YOUR ANSWER HERE. Blank-plot (hint: I gave you the answer in the chunk name)

(Notice how I used `fig.width` and `fig.height` in the chunk options. You can click on the little gear icon in the far left of the chunk to change other options.)

**2: How many rows are in `mpg`? How many columns?**

```{r mpg-details}
nrow(mpg)
ncol(mpg)

# Or
dim(mpg)

# Or
mpg
```

**3: What does the `drv` variable describe? Read the help for `?mpg` to find out.**

TYPE YOUR ANSWER HERE. There are 234 rows and 11 columns. drv means the following:
f = front-wheel drive, r = rear wheel drive, 4 = 4wd

**4: Make a scatterplot of `hwy` vs `cyl`.**

```{r hwy-cyl-scatterplot}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = cyl, y = hwy))
```

**5: What happens if you make a scatterplot of `class` vs `drv`? Why is the plot not useful?**

```{r class-drv-scatterplot}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = class, y = drv))
```

TYPE YOUR ANSWER HERE. They both are the same type overlapping one on another. In other words, drv and class are categorical 



## 3.3.1
**1: What’s gone wrong with this code? Why are the points not blue?**
```{r disp-hwy-geompoint}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```

It is simple as color argument should not be written inside aes()

```{r disp-hwy-geompoint, correct version}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```
**Which variables in mpg are categorical? Which variables are continuous? (Hint: type ?mpg to read the documentation for the dataset). How can you see this information when you run mpg?**
```{r view-data}
mpg
```
-Categorical: manufacturer, model, trans, drv,fl, class
-Continuous: hwy, cty, cyl, displ

**Map a continuous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs. continuous variables?**
```{r color, size, shape for continuous}
ggplot(data = mpg) + 
geom_point(mapping = aes(x = displ, y = hwy, color = cyl))
ggplot(data = mpg) + 
geom_point(mapping = aes(x = displ, y = hwy, size = cyl))
ggplot(data = mpg) + 
geom_point(mapping = aes(x = displ, y = hwy, shape = cyl))

```
```{r color, size, shape for categorical}
ggplot(data = mpg) + 
geom_point(mapping = aes(x = displ, y = hwy, color = drv))
ggplot(data = mpg) + 
geom_point(mapping = aes(x = displ, y = hwy, size = drv))
ggplot(data = mpg) + 
geom_point(mapping = aes(x = displ, y = hwy, shape = drv))

```
-Using color function for continous data gives us one color but in different spectrum (palette) while in categorical color is distinct for each category. Also, size is not  preffered for categorical. Also, continous data cant take shape function.

**What happens if you map the same variable to multiple aesthetics?**
```{r same variable to multiple aesthetics }
ggplot(data = mpg) + 
geom_point(mapping = aes(x = displ, y = hwy, color =drv, shape=drv))
```

-As we can see both functions are used.

**What does the stroke aesthetic do? What shapes does it work with? (Hint: use ?geom_point)**
```{r stroke }
?geom_point
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy), stroke = 2, shape=21)
```
For shapes that have a border (like 21), we can colour the inside and
outside separately. Use the stroke aesthetic to modify the width of the border

**What happens if you map an aesthetic to something other than a variable name, like aes(colour = displ < 5)? Note, you’ll also need to specify x and y.**
```{r special aesthetic case }
ggplot(data=mpg)+
geom_point(mapping=aes(x=displ, y=hwy, colour = displ < 5))
```

## 3.5.1

**What happens if you facet on a continuous variable?**
```{r facet }
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = drv, y = cyl)) +
  facet_wrap(~ hwy)
```
It does not bring any logical sense. 

**What do the empty cells in plot with facet_grid(drv ~ cyl) mean? How do they relate to this plot?**
```{r compare two plots}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = drv, y = cyl))

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = drv, y = cyl)) + 
  facet_grid(drv ~ cyl)
```
To facet our plot on the combination of two variables, add facet_grid() to our plot call. The first argument of facet_grid() is also a formula. This time the formula should contain two variable names separated by a ~. Both plots are similar.

**What plots does the following code make? What does . do?**
```{r what does . make}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(. ~ cyl)
```
facet_grid(.~variable) will return facets equal to the levels of variable distributed horizontally.

facet_grid(variable~.) will return facets equal to the levels of variable distributed vertically.

**What are the advantages to using faceting instead of the colour aesthetic? What are the disadvantages? How might the balance change if you had a larger dataset?**
```{r pros and cons of facet}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```
-faceting helps to identify trend for each class seperatelly while colour aesthetic is good for visualizing overall class pattern 
-when dataset size is large  that will not make much sense as more data will sit on one another making visualizing less transperent.

**Read ?facet_wrap. What does nrow do? What does ncol do? What other options control the layout of the individual panels? Why doesn’t facet_grid() have nrow and ncol arguments?**

nrow sets how many rows the faceted plot will have.
ncol sets how many columns the faceted plot will have.
facet_grid() doesn’t need nrow and ncol because they are already defined by faceting variables. 
**When using facet_grid() you should usually put the variable with more unique levels in the columns. Why?**
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid( trans~ drv)
```
As can be seen, the understanding of graph will be less visible. Graph compressed and therefore less appropriate to visualize.

## 3.6.1
**Geometric objects**
What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?
Line chart - geom_line(), Boxplot - geom_boxplot(), Histogram - geom_histogram(), Area chart - geom_area()

```{r, geom objects}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point() + 
  geom_smooth(se = FALSE)
```
**What does show.legend = FALSE do? What happens if you remove it?**
-It removes the legend. aesthetics are drawn but legend is omitted

**What does the se argument to geom_smooth() do?**
-It draws the confidence intervals in the smoothed lines.

**Will these two graphs look different? Why/why not?**

```{r, compare difference}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```
The former code will produce the same plot as the the latter code:



## 3.8.1
